A Shared Subspace Model for Multi-Group Covariance Estimation

### 2016-07-06 paper/mscov.tex
Minor edits from proofread

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow and a
    graduate of the Department of Statistics at Harvard University
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter Hoff is a
    Professor of Statistics at The University of Washington
    (\href{mailto:pdhoff@uw.edu}{pdfhoff@uw.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in Data-Intensive Discovery and the Moore/Sloan Data Science Environments Project at the University of Washington.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington) and Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) for sharing data and ideas
 which contributed to the framing of the ideas this paper.}}
\author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle

@@

  We develop a model-based method for evaluating heterogeneity among
  several $p\times p$ covariance matrices in the big p, small n setting.
  This is done by assuming a spiked covariance model for each group
  and sharing information about the space spanned by the group-level
  eigenvectors.  We use an empirical Bayes method to identify a low-dimensional
  subspace which explains variation across all groups and use an MCMC
  algorithm to estimate the posterior uncertainty of eigenvectors and
  eigenvalues on this subspace.  The implementation and utility of our
  model is illustrated with analyses of high-dimensional multivariate
  gene expression and metabolomics data.
\vfill
\noindent\textbf{Keywords:} covariance estimation; spiked covariance
model; Stiefel manifold; Grassmann manifold; big p, small n;
high-dimensional data; empirical Bayes.
\end{abstract}

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS). This approach, which is a special case of a bilinear factor
model, involves projecting the data onto a lower dimensional space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  A
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage.

In this paper we propose a multi-group covariance estimation model by
sharing information about the subspace spanned by group-level
eigenvectors.  The shared subspace assumption can be used to improve
estimates and facilitate the interpretation of differences between
covariance matrices across groups.  For each group, we assume ``the
spiked covariance model'' (also known as the ``partial isotropy
model''), a well studied variant of the factor model
\citep{Mardia1980, Johnstone2001}.  In Section \ref{sec:shared} we
briefly review the behavior of spiked covariance models for estimating
a single covariance matrix and then introduce our extension to the
multi-group setting.  In Section \ref{sec:inference} we describe an
empirical Bayes algorithm for inferring the shared subspace and
estimating the posterior distribution of the covariance matrices of
the data projected onto this subspace.  In Section
\ref{sec:simulation} we investigate the behavior of this class of
models in simulation and demonstrate how the shared subspace
assumption is widely applicable, even when there is little similarity
in the covariance matrices across groups.  In particular, shared
subspace models include those in which there is no information shared
across groups because in the worst case we can assume the subspace
spanned by the group-level eigenvectors is simply the full
$p$-dimensional space.  We then demonstrate the utility of the model
in an analysis of gene expression data from juvenile leukemia patients
in Section \ref{sec:app}.  Despite the large feature size ($p > 3000$)
and small sample size ($n < 100$ per group), we identify interpretable
similarities and differences in gene covariances on a low dimensional
subspace.  Finally, we conclude with a brief exploration of
heterogeneity in metabolomic data from a study on aging in the fly
species \textit{Drosophila melanogaster}.

@@

Suppose $S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(S | \Sigma, n) \propto l(\Sigma: S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1}  S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
studied by \citet{Johnstone2001} and others assumes that
\begin{equation} 
\Sigma = \sigma^2 ( U  \Lambda  U^T  + I )
\label{eqn:spiked}
\end{equation}
%
\noindent where for $r \ll p$, $\Lambda$ is an $r\times r$ diagonal
matrix and $U \in \mathcal V_{r,p}$.  $\mathcal V_{r,p}$ is the
Stiefel manifold, which consists of the set of $p \times r$
orthonormal matrices in $\mathbb{R}^p$, so that $U^TU =
I$.  The spiked covariance formulation is appealing because it explicitly
partitions the covariance matrix into a tractable low rank ``signal''
and isotropic ``noise''.

@@

Note that covariance estimators of this form are are equivariant with
respect to rotations and scale changes.  The situation should be
different, however, when we are interested in estimating multiple
covariance matrices from distinct, but related groups.  Here,
group-level equivariance to rotations is an unreasonable assumption;
both eigenvalue \emph{and} eigenvector shrinkage can play an important
role in improving covariance estimates.  Consider multi-group
covariance estimation based on $K$ matrices, $Y_1, ..., Y_K$, where
$Y_k$ is assumed to be an $n_k$ by $p$ matrix of mean-zero normal
data, typically with $n_k \ll p$.  Then, $Y_k^TY_k = S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators that share information across
groups, that is, where the estimator $\hat{\Sigma}_k$ may depend on
data from all groups.  To this end, we extend the spiked covariance
model to grouped data, by assuming that the anisotropic variability
from each group occurs on a common low dimensional subspace.
Specifically, we posit that the covariance matrix for each group can
be written as
\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI,
\label{eqn:sspsi}
\end{equation}

@@

In this section we outline an empirical Bayesian approach for
estimating a low-dimensional shared subspace and the covariance
matrices of the data projected onto this space. As we highlight in
Section \ref{sec:simulation}, if the spiked covariance model holds for
each group individually, then the shared subspace assumption also
holds because at worst we can take the shared subspace ot be the full
$p$-dimensional space. In practice, we can usually identify a shared
subspace of dimension $s \ll p$ that preserves most of the variation
in the data.  Our primary objective is to identify the ``best'' shared
subspace of fixed dimension $s < p$.  In Section \ref{sec:em} we
describe an expectation maximization algorithm for estimating the
maximum marginal likelihood of the shared subspace, $VV^T$.  This
approach is computationally tractable for high-dimensional datasets.
Given an inferred subspace, we then seek estimators for the covariance
matrices of the data projected onto this space.  Because seemingly
large differences in the point estimates of covariance matrices across
groups may not actually reflect statistically significant differences,
in Section \ref{sec:bayes} we also describe a Gibbs sampler that can
be used to generate estimates of the projected covariance matrices,
$\Psi_k$, and and their associated uncertainty.  Later, we discuss
strategies for infering an appropriate value for $s$ and explore how
shared subspace models can be used for exploratory data analysis by
visualizing covariance heterogeneity on two or three dimensional
subspaces.

@@

In this section we describe a maximum marginal likelihood procedure for
estimating the shared subspace, $VV^T$ based on the
expectation-maximization algorithm.  The full likelihood
for the shared subspace model can be written as

@@

We propose an expectation-maximization (EM) algorithm to estimate the
relevant quantities.  In particular we maximize the marginal
likelihood of $V$, treating $M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as
the ``missing'' parameters to be imputed.  We assume independent
Jeffreys' prior distributions for both $\sigma_k^2$ and
$M_k$.  Jeffreys' prior for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it
can easily be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-1/2(M_k^{-1}V^TS_kV)) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(-\frac{1}{2\sigma^2_k} (I-VV^T)S_k\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\frac{1}{2}\text{tr}((I-VV^T)S_k)$) distribution.  Based on these
results the following EM algorithm can be used:

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T S_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-s)}{tr((I-V_{(t-1)}V_{(t-1)}^T)S_k)}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{\text{arg } \text{max}}  \sum_k tr\left(-VE[M^{-1}|V_{(t-1)}]^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)S_k/2)$ 
\end{enumerate}

For the second step (``M-step''), we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by dimension of the shared subspace, $s$, not the number of
features $p$.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

If $V$ is a basis for a shared subspace, then for each group $k$, most
of the non-isotropic variation in $Y_k$ should be preserved when projecting the data
to this space.  To characterize the extent to which this is true for
different groups, we propose a simple estimator for the proportion of
``signal'' variance that lies on any given subspace.  Specifically,
for an estimated subspace $\hat{V}$, we use the following statistic
for the ratio of the sum of the first $s$ eigenvalues of
$\hat{V}^T \Sigma_k \hat{V}$ to the sum of the first $s$
eigenvalues of $\Sigma_k$:
%
\begin{equation}
\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2) = \frac{||Y_k\hat{V}||_F/n_k}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F/n_k - \hat{\sigma}_k^2ps/n_k}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||\cdot||_F$ is the Frobenius norm. Our motivation for
this statistic is the fact that if
$\Sigma_k = \hat{V}\Psi_k\hat{V}^T + \sigma_k^2I$ then
$\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2) \approx 1$.  We demonstrate
this in an asymptotic regime where $s$ is fixed, $p$,
$n_k \rightarrow \infty$ and $p/n_k = \alpha_k$ constant.

First, since $s$ is fixed and $n_k$ is growing, the numerator,
$||Y_k\hat{V}||_F/n_k$, is a consistent estimator for the sum of the
eigenvalues of $\hat{V}^T\Sigma_k\hat{V}$.  The denominator,
$\underset{\widetilde{V} \in \mathcal{V}_{p,
    s}}{\text{max}}||Y_k\widetilde{V}||_F/n_k$
is equivalent to the sum of the first $s$ eigenvalues of the sample
covariance matrix $S_k/n_k$.  With $\hat{\lambda}^{(k)}_i$ the
$i$-th eigenvalue of $S_k/n_k$, \citet{Baik2006} demonstrated that
asymptotically as $p, n_k \rightarrow \infty$ with $p/n_k = \alpha_k$
fixed
%
\begin{eqnarray}
\nonumber \hat{\lambda}^{(k)}_i &\rightarrow& \lambda^{(k)}_i\left(1 +
                                    \frac{\sigma_k^2\alpha_k}{\lambda^{(k)}_i
                                    - 1}\right)\\
& \approx& \lambda^{(k)} + \sigma^2_k\frac{p}{n_k}
\end{eqnarray}
%
\noindent where the approximation in the second line is due to the
fact that $\lambda_k/(\lambda_k - 1) \approx 1$ for $\lambda_k$
large. Thus,
$\sum_i^s \hat{\lambda}^{(k)}_i \approx \sum_i^s \lambda^{(k)}_i +
\sigma_k^2ps/n_k$
and $\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F/n_k - \hat{\sigma}_k^2ps/n_k$ is a reasonable approximation to the sum of the
first $s$ eigenvalues of $\Sigma_k$.

As a consequence, the proposed statistic will be
close to one for all groups when $\hat{V}\hat{V}^T = VV^T$ and smaller
if not.  The metric can provides a useful indicator for which groups
can be reasonably compared on a given subspace and which groups
cannot.  When $\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2)$ is small for
some groups, it may suggest that the rank, $s$, of the inferred
subspace needs to be larger to capture the variation in all groups. To
demonstrate this, in Section \ref{sec:simulation}, we compute these
statistics for inferred subspaces of different dimensions on a single
dataset. In Section \ref{sec:app}, we plot the estimates for subspaces
inferred with real biological data.

@@

There are many different ways in which we could choose to parameterize
$\Psi_k$.  Building on recent interest in the spiked covariance model
\citep{Donoho2013, Paul2007} we propose a tractable MCMC algorithm by
specifying priors on the eigenvalues and eigenvectors of $\Psi_k$.  By
modeling the eigenstructure, we can now view each covariance
$\Sigma_k$ in terms of the original spiked principle components model
(\ref{eqn:spiked}).  Equation \ref{eqn:sspsi}, written as a function of
$V$, becomes
%
\begin{align}
\nonumber \Psi_k &= O_k\Lambda_KO_k^T\\
\Sigma_k &= V\Psi_kV^T + \sigma^2_kI.
\label{eqn:ss}
\end{align}
%
\noindent Here, we allow $\Psi_k$ to be of rank $r \leq s$ dimensional
covariance matrix on the $s$-dimensional subspace.  Thus, $\Lambda_k$
is an $r \times r$ diagonal matrix of eigenvalues, and
$O_k \in \mathcal{V}_{s,r}$ is the matrix of eigenvectors of $\Psi_k$.
For any individual group, this corresponds to the original spiked PCA
model (Equation \ref{eqn:spiked}) with
$U_k = VO_k \in \mathcal{V}_{p, r}$.  This generalization is helpful
because it enables us to independently specify a subspace common to
all groups and the possibly lower rank features on this space that are
specific to individual groups.  Although our model is most useful when
the covariance matrices are related across groups, we can also use
this formulation to specify models for multiple unrelated spiked
covariance models.  We explore this in detail in Section
\ref{sec:simulation}.  In Section \ref{sec:app} we introduce a shared
subspace model with additional structure on the eigenvectors and
eigenvalues of $\Psi_k$ to facilitate interpretation of covariance
heterogeneity on a two-dimensional subspace.

@@

From Equation \ref{eqn:sslik} it is clear that the the inverse-gamma
class of prior distributions is conjugate for $\sigma_k^2$.  We chose a
default prior distribution for $\sigma^2$ that is equivariant with
respect to scale changes.  Specifically, we use Jeffreys' prior, an
improper prior with density $p(\sigma^2) \propto 1/\sigma^2 $.  Under
this prior, straightforward calculations show that the full
conditional distribution of $\sigma_k^2$ is
inverse-gamma$( n_k p/2 , \tr[S_k( I - U_k\Omega_k
U_k)/2])$, where $U_k = VO_k$.

\paragraph{Conditional distribution of $O_k$:} Given the likelihood
from Equation \ref{eqn:sslik}, it is easy to show that the class of
Bingham distributions are conjugate for $O_k$ \citep{Hoff2009,
  Hoff2012}.  Again, invariance considerations lead us to use a
rotationally invariant uniform probability measure on
$\mathcal V_{s,p}$.  Under this uniform prior, the full conditional
distribution of $O_k$ has a density proportional to the
likelihood, so
%
\begin{align}
\label{lik_vo}
 p(O_k | \sigma^2_k, U_k, \Omega_k) & \propto \etr(\Omega_kO^T_kV^T[S_k/(2\sigma^2_k)]VO_k).
\end{align}
%
\noindent This is a Bingham$(\Omega, \hat{V}^T S_k \hat{V}/(2\sigma^2))$
distribution on $\mathcal V_{s, r}$ \citep{Khatri1977}. A
Gibbs sampler to simulate from this distribution is given in
\citet{Hoff2012}.

>>>>

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow and a
    graduate of the Department of Statistics at Harvard University
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter D. Hoff
    is a Professor in the Departments of Statistics, Biostatistics and
    the Center for Statistics and the Social Sciences at the
    University of Washington
    (\href{mailto:pdhoff@uw.edu}{pdfhoff@uw.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in Data-Intensive Discovery and the Moore/Sloan Data Science Environments Project at the University of Washington.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington) and Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) for sharing data and ideas
 which contributed to framing of this paper.}}
\author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle

@@

  We develop a model-based method for evaluating heterogeneity among
  several $p\times p$ covariance matrices in the big $p$, small $n$ setting.
  This is done by assuming a spiked covariance model for each group
  and sharing information about the space spanned by the group-level
  eigenvectors.  We use an empirical Bayes method to identify a low-dimensional
  subspace which explains variation across all groups and use an MCMC
  algorithm to estimate the posterior uncertainty of eigenvectors and
  eigenvalues on this subspace.  The implementation and utility of our
  model is illustrated with analyses of high-dimensional multivariate
  gene expression and metabolomics data.
\vfill
\noindent\textbf{Keywords:} covariance estimation; spiked covariance
model; Stiefel manifold; Grassmann manifold; big p, small n;
high-dimensional data; empirical Bayes.
\end{abstract}

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS) \citep{Wold2001}. This approach, which is a special case of a bilinear factor
model, involves projecting the data onto a lower dimensional space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  A
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage.

In this paper we propose a multi-group covariance estimation model by
sharing information about the subspace spanned by group-level
eigenvectors.  The shared subspace assumption can be used to improve
estimates and facilitate the interpretation of differences between
covariance matrices across groups.  For each group, we assume ``the
spiked covariance model'' (also known as the ``partial isotropy
model''), a well studied variant of the factor model
\citep{Mardia1980, Johnstone2001}.  In Section \ref{sec:shared} we
briefly review the behavior of spiked covariance models for estimating
a single covariance matrix and then introduce our extension to the
multi-group setting.  In Section \ref{sec:inference} we describe an
empirical Bayes algorithm for inferring the shared subspace and
estimating the posterior distribution of the covariance matrices of
the data projected onto this subspace.

In Section \ref{sec:simulation} we investigate the behavior of this
class of models in simulation and demonstrate how the shared subspace
assumption is widely applicable, even when there is little similarity
in the covariance matrices across groups.  In particular, shared
subspace models include those in which there is no information shared
across groups because in the worst case we can assume the subspace
spanned by the group-level eigenvectors is simply the full
$p$-dimensional space.  In Section \ref{sec:app} we demonstrate the
utility of a shared subspace model in an analysis of gene expression
data from juvenile leukemia patients .  Despite the large feature size
($p > 3000$) and small sample size ($n < 100$ per group), we identify
interpretable similarities and differences in gene covariances on a
low dimensional subspace.  Finally, we conclude with a brief
exploration of heterogeneity in metabolomic data from a study on aging
in the fly species \textit{Drosophila melanogaster}.

@@

Suppose $S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(S | \Sigma, n) \propto l(\Sigma: S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1}  S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
studied by \citet{Johnstone2001} and others assumes that
\begin{equation} 
\Sigma = \sigma^2 ( U  \Lambda  U^T  + I )
\label{eqn:spiked}
\end{equation}
%
\noindent where for $r \ll p$, $\Lambda$ is an $r\times r$ diagonal
matrix and $U \in \mathcal V_{p,r}$.  $\mathcal V_{p,r}$ is the
Stiefel manifold, which consists of the set of $p \times r$
orthonormal matrices in $\mathbb{R}^p$, so that $U^TU =
I$.  The spiked covariance formulation is appealing because it explicitly
partitions the covariance matrix into a tractable low rank ``signal''
and isotropic ``noise''.

@@

Note that covariance estimators of this form are are equivariant with
respect to rotations and scale changes.  The situation should be
different, however, when we are interested in estimating multiple
covariance matrices from distinct, but related groups.  Here,
group-level equivariance to rotations is an unreasonable assumption;
both eigenvalue \emph{and} eigenvector shrinkage can play an important
role in improving covariance estimates.  Consider multi-group
covariance estimation based on $K$ matrices, $Y_1, ..., Y_K$, where
$Y_k$ is assumed to be an $n_k$ by $p$ matrix of mean-zero normal
data, typically with $n_k \ll p$.  Then, $Y_k^TY_k = S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators of each covariance matrix,
$\hat{\Sigma}_k$, that may depend on data from all groups.  To this
end, we extend the spiked covariance model to grouped data, by
assuming that the anisotropic variability from each group occurs on a
common low dimensional subspace.  Specifically, we posit that the
covariance matrix for each group can be written as
\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI,
\label{eqn:sspsi}
\end{equation}

@@

In this section we outline an empirical Bayesian approach for
estimating a low-dimensional shared subspace and the covariance
matrices of the data projected onto this space. As we highlight in
Section \ref{sec:simulation}, if the spiked covariance model holds for
each group individually, then the shared subspace assumption also
holds because at worst we can take the shared subspace to be the full
$p$-dimensional space. In practice, we can usually identify a shared
subspace of dimension $s \ll p$ that preserves most of the variation
in the data.  Our primary objective is to identify the ``best'' shared
subspace of fixed dimension $s < p$.  In Section \ref{sec:em} we
describe an expectation maximization algorithm for estimating the
maximum marginal likelihood of the shared subspace, $VV^T$.  This
approach is computationally tractable for high-dimensional datasets.
Given an inferred subspace, we then seek estimators for the covariance
matrices of the data projected onto this space.  Because seemingly
large differences in the point estimates of covariance matrices across
groups may not actually reflect statistically significant differences,
in Section \ref{sec:bayes} we also describe a Gibbs sampler that can
be used to generate estimates of the projected covariance matrices,
$\Psi_k$, and and their associated uncertainty.  Later, we discuss
strategies for infering an appropriate value for $s$ and explore how
shared subspace models can be used for exploratory data analysis by
visualizing covariance heterogeneity on two or three dimensional
subspaces.

@@

In this section we describe a maximum marginal likelihood procedure for
estimating the shared subspace, $VV^T$, based on the
expectation-maximization (EM) algorithm.  The full likelihood
for the shared subspace model can be written as

@@

We maximize the marginal likelihood of $V$ with an EM algorithm, where
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as the considered the
``missing'' parameters to be imputed.  We assume independent Jeffreys'
prior distributions for both $\sigma_k^2$ and $M_k$.  Jeffreys' prior
for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it can easily
be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-1/2(M_k^{-1}V^TS_kV)) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(-\frac{1}{2\sigma^2_k} (I-VV^T)S_k\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\frac{1}{2}\text{tr}[(I-VV^T)S_k]$) distribution.  Based on these
results, we complete the following steps for each iteration, $t$, of
the EM algorithm until convergence:

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T S_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-s)}{\text{tr}[(I-V_{(t-1)}V_{(t-1)}^T)S_k]}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{\text{arg } \text{max}}  \sum_k \text{tr}\left(-VE[M^{-1}|V_{(t-1)}]^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)S_k/2)$ 
\end{enumerate}

For the second step (``M-step''), we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by the dimension of the shared subspace not the number of
features.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

If $V$ is a basis for a shared subspace, then for each group, most
of the non-isotropic variation in $Y_k$ should be preserved when
projecting the data to this space.  To characterize the extent to
which this is true for different groups, we propose a simple estimator
for the proportion of ``signal'' variance that lies on a given
subspace.  Specifically, we use the following statistic for the ratio
of the sum of the first $s$ eigenvalues of
$V^T \Sigma_k V$ to the sum of the first $s$ eigenvalues
of $\Sigma_k$:
%
\begin{equation}
\gamma(Y_k: V, \sigma_k^2) = \frac{||Y_kV||_F/n_k}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F/n_k - \sigma_k^2ps/n_k}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||\cdot||_F$ is the Frobenius norm. Our motivation for
this statistic is the fact that if
$\Sigma_k = V\Psi_kV^T + \sigma_k^2I$ then
$\gamma(Y_k: V, \sigma_k^2) \approx 1$.  We establish
this in an asymptotic regime where $s$ is fixed, $p$,
$n_k \rightarrow \infty$ and $p/n_k = \alpha_k$ constant.

First, since $s$ is fixed and $n_k$ is growing, the numerator,
$||Y_kV||_F/n_k$, is a consistent estimator for the sum of the
eigenvalues of $V^T\Sigma_kV$.  The denominator,
$\underset{\widetilde{V} \in \mathcal{V}_{p,
    s}}{\text{max}}||Y_k\widetilde{V}||_F/n_k$
is equivalent to the sum of the first $s$ eigenvalues of the sample
covariance matrix $S_k/n_k$.  With $\hat{\lambda}^{(k)}_i$ the
$i$-th eigenvalue of $S_k/n_k$, \citet{Baik2006} demonstrated that
asymptotically as $p, n_k \rightarrow \infty$ with $p/n_k = \alpha_k$
fixed
%
\begin{eqnarray}
\nonumber \hat{\lambda}^{(k)}_i &\rightarrow& \lambda^{(k)}_i\left(1 +
                                    \frac{\sigma_k^2\alpha_k}{\lambda^{(k)}_i
                                    - 1}\right)\\
& \approx& \lambda^{(k)} + \sigma^2_k\frac{p}{n_k}
\label{eqn:lamAsymp}
\end{eqnarray}
%
\noindent where the approximation in the second line is due to the
fact that $\lambda_k/(\lambda_k - 1) \approx 1$ for $\lambda_k$ large.
Thus,
$\sum_i^s \hat{\lambda}^{(k)}_i \approx \sum_i^s \lambda^{(k)}_i +
\sigma_k^2ps/n_k$
and
$\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
||Y_k\widetilde{V}||_F/n_k - \hat{\sigma}_k^2ps/n_k$
is a reasonable approximation to the sum of the first $s$ eigenvalues
of $\Sigma_k$.  Alternatively, when $\lambda_k$ is small, a better
approximation may be obtained by solving the quadratic equation
implied by the first line of Equation \ref{eqn:lamAsymp}.

As a consequence, the proposed statistic will be close to one for all
groups when $VV^T$ is a shared subspace for the data and smaller if
not.  The metric provides a useful indicator for which groups can be
reasonably compared on a given subspace and which groups cannot.  In
practice, we estimate a shared subspace, $\hat{V}$, and the isotropic
variances, $\hat{\sigma}_k^2$, using EM and compute the plug-in
estimate $\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2)$.  When this
statistic is small for some groups, it may suggest that the rank, $s$,
of the inferred subspace needs to be larger to capture the variation
in all groups. We investigate this in Section \ref{sec:simulation}, by
computing the goodness of fit statistic for inferred subspaces of
different dimensions on a single dataset. In Section \ref{sec:app}, we
compute the estimates for subspaces inferred with real biological
data.

@@

There are many different ways in which we could choose to parameterize
$\Psi_k$.  Building on recent interest in the spiked covariance model
\citep{Donoho2013, Paul2007} we propose a tractable MCMC algorithm by
specifying priors on the eigenvalues and eigenvectors of $\Psi_k$.  By
modeling the eigenstructure, we can now view each covariance
$\Sigma_k$ in terms of the original spiked principle components model.  Equation \ref{eqn:sspsi}, written as a function of
$V$, becomes
%
\begin{align}
\nonumber \Psi_k &= O_k\Lambda_KO_k^T\\
\Sigma_k &= V\Psi_kV^T + \sigma^2_kI.
\label{eqn:ss}
\end{align}
%
\noindent Here, we allow $\Psi_k$ to be of rank $r \leq s$ dimensional
covariance matrix on the $s$-dimensional subspace.  Thus, $\Lambda_k$
is an $r \times r$ diagonal matrix of eigenvalues, and
$O_k \in \mathcal{V}_{s,r}$ is the matrix of eigenvectors of $\Psi_k$.
For any individual group, this corresponds to the original spiked PCA
model (Equation \ref{eqn:spiked}) with
$U_k = VO_k \in \mathcal{V}_{p, r}$.  Differentiating the ranks $r$
and $s$ is helpful because it enables us to independently specify a
subspace common to all groups and the possibly lower rank features on
this space that are specific to individual groups.  Although our model
is most useful when the covariance matrices are related across groups,
we can also use this formulation to specify models for multiple
unrelated spiked covariance models.  We explore this in detail in
Section \ref{sec:simulation}.  In Section \ref{sec:app} we introduce a
shared subspace model with additional structure on the eigenvectors
and eigenvalues of $\Psi_k$ to facilitate interpretation of covariance
heterogeneity on a two-dimensional subspace.

@@

From Equation \ref{eqn:sslik} it is clear that the the inverse-gamma
class of prior distributions is conjugate for $\sigma_k^2$.  We chose a
default prior distribution for $\sigma^2_k$ that is equivariant with
respect to scale changes.  Specifically, we use Jeffreys' prior, an
improper prior with density $p(\sigma^2_k) \propto 1/\sigma^2_k $.  Under
this prior, straightforward calculations show that the full
conditional distribution of $\sigma_k^2$ is
inverse-gamma$( n_k p/2 , \tr[S_k( I - U_k\Omega_k
U_k)/2])$, where $U_k = VO_k$.

\paragraph{Conditional distribution of $O_k$:} Given the likelihood
from Equation \ref{eqn:sslik}, it is easy to show that the class of
Bingham distributions are conjugate for $O_k$ \citep{Hoff2009,
  Hoff2012}.  Again, invariance considerations lead us to use a
rotationally invariant uniform probability measure on
$\mathcal V_{s,p}$.  Under this uniform prior, the full conditional
distribution of $O_k$ has a density proportional to the
likelihood, so
\begin{align}
\label{lik_vo}
 p(O_k | \sigma^2_k, U_k, \Omega_k) & \propto \etr(\Omega_kO^T_kV^T[S_k/(2\sigma^2_k)]VO_k).
\end{align}
%
\noindent This is a Bingham$(\Omega, V^T S_k V/(2\sigma^2))$
distribution on $\mathcal V_{s, r}$ \citep{Khatri1977}. A
Gibbs sampler to simulate from this distribution is given in
\citet{Hoff2012}.

Together, the prior for $\sigma_k^2$ and $O_k$ leads to conditional
(on $V$) Bayes estimators $\hat \Sigma(V^T S_k V)$ that are
equivariant with respect to scale changes and rotations on the
subspace spanned by $V$, so
that $\hat \Sigma(a W V^T S_k V W^T) = a W \hat\Sigma(V^T
S_k V)  W$
for all $a>0$ and $ W\in \mathcal O_{s}$ (assuming an invariant
loss function). Interestingly, if $\Omega_k$ were known (which it is
not), then for a given invariant loss function the Bayes estimator
under this prior minimizes the (frequentist) risk among all
equivariant estimators \citep{Eaton1989}.
